Locality and Effect Reflection

### 2025-10-06 paper.tex
More changes

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
  $\glob \tau$, $\yoneda{\Sigma}$ is a type representing how to inject
  terms of $\Sigma$ into the current ambient monad, and $\glob$ is a
  comonadic \emph{global modality} used to track which values do not
  depend on the ambient monad. $\ind{\Sigma}$ is sometimes called the
  \emph{freer monad} of $\Sigma$. This isomorphism emerges naturally
  from a denotational semantics for algebraic effects in terms of
  presheaves over bounded Lawvere theories.

@@

  let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
    let rec loop (s : int) : 'a Term(State).t @ local -> 'a  = function
      | Return x -> x
      | Op(Get, k) -> loop s (k s)
      | Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
  let get (h : Handler(State).t @ local) : int =
    State_eff.perform Get h

@@

\subsection{Denotational semantics}

This design emerges naturally from a denotational semantics where types
are interpreted as presheaves over \emph{bounded Lawvere theories} -- a
categorical framework that makes the connection between locality and
effects precise.

@@

The key constructs of effect reflection are then defined as follows:
\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebras.

@@

\item The effect reflection isomorphism becomes a simple consequence of
  the Yoneda lemma and some properties of the sums of algebraic
  theories.
\end{itemize}

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors. We start with a simple module type
\lstinline[style=oxcaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=oxcaml]
  module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
and \lstinline[style=oxcaml]{Handler} types.
\begin{lstlisting}[style=oxcaml]
  module Term (O : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
  end

  module Handler(O : Op) : sig
    type t
  end
\end{lstlisting}
These types directly implement the types from our core isomorphism:
\lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
$\ind{\Sigma}(\glob \tau)$, while
\lstinline[style=oxcaml]{Handler(Sigma).t} corresponds to
$\yoneda{\Sigma}$

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
represents the algebraic effect handlers that are handling the current
computation.

@@

@@

    val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
  end
\end{lstlisting}
\lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_local} implement the left-to-right
direction. We need two separate functions because OxCaml lacks any form
of mode polymorphism.

\lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
direction but makes for a more convenient primitive in
practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
individual operations as opposed to full terms. Using
\lstinline[style=oxcaml]{perform} we can implement reflection of global
terms as:
\begin{lstlisting}[style=oxcaml]
  let rec reflect t h =
    match t with
    | Return x -> x
    | Op(op, k) -> reflect (k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
  val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
\end{lstlisting}
Reflection of local terms can be defined similarly.

@@

  let handle_state i f =
    let rec loop (s : int) = function
      | Return x -> x
      | Op(Get, k) -> loop s (k s)
      | Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=oxcaml]
  type handler = Handler(State).t

@@

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented directly using mutable state
and locality:
\begin{lstlisting}[style=oxcaml]
  type handler = int ref

  let handle_state i f = f (ref i)
  let get r = !r
  let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state monad. In
some sense, the improved reasoning ability of the state monad over
ordinary mutable state is precisely captured by the notion of locality.

This transformation from effect reflection to local mutable state works
for any \emph{tail-resumptive} effect handler.

@@

  let handle_gen f =
    let rec loop = function
      | Return () -> Finished
      | Op(Gen i, k) -> More(i, fun () -> loop (k ()))
    in
    loop (Gen_eff.reify f)

\end{lstlisting}

@@

  let handle_await f =
    let rec loop = function
      | Return x -> Deferred.return x
      | Op(Await d, k) ->
          Deferred.bind d (fun x -> loop (k x))
    in
    loop (Await_eff.reify f)

@@

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ah ->
    handle_state i (fun sh ->
      List.iter (fun fl ->
        let s = await ah (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh)) files;
      get sh))
\end{lstlisting}

@@

We could change the definition of \lstinline[style=oxcaml]{handle_gen} to
use \lstinline[style=oxcaml]{reify_local} instead of
\lstinline[style=oxcaml]{reify}, but then we would lose the property that
the generator we produce can be passed around freely. Instead we need to
build a handler that can forward the \lstinline[style=oxcaml]{Await}
operation to different effect handlers at different points in the
program's execution.

@@

Composing handlers, like \lstinline[style=oxcaml]{handle_gen}, that use
\lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_local} is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=oxcaml]{Await} operations, so whenever we force part of
the stream we'll need to provide a handler for those
\lstinline[style=oxcaml]{Await} operations again.

@@

  let handle_gen_in_await f ah =
    let rec loop ah = function
      | Return () -> Finished
      | Op(Left op, k) -> loop ah (k (perform ah op))
      | Op(Right(Gen i), k) -> More(i, fun ah -> loop ah (k ()))
    in
    loop ah (Eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))
\end{lstlisting}

The need for effect forwarding is a general short-coming with lexical
handlers that applies to all such systems. Examples like this
fundamentally require that performing an \lstinline[style=oxcaml]{Await}
operation to does work proportional to the number of handlers it is
forwarded through. Lexical handlers do a constant amount of work to
perform operations and so cannot handle such examples
directly. Non-lexical effect handlers *always* do work proportional to
the number of nested handlers, allowing them to support such examples
directly but paying a higher cost for the cases that do not need
forwarding.

@@

\begin{figure}
\begin{align*}
\fst{(v_1, v_2)} &\reducestov v_1 \\
\snd{(v_1, v_2)} &\reducestov v_2 \\
\gunbox{x}{\gbox{v_1}}{v_2} &\reducestov \sbst{v_2}{x}{v_1} \\
\grun{\return{v}} &\reducestov v \\
\letv{x}{\return{v}}{c} &\reducestov \sbst{c}{x}{v} \\
\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestov \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
\app{(\abs{x}{\tau}{c})}{v} &\reducestoc \sbst{c}{x}{v} \\
\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v} \\
\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v} \\
\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducestoc \sbst{c}{x}{v} \\
\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
  \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
\reify{\Sigma}{x}{\return{v}} &\reducestoc \return{\ret{\Sigma}{v}} \\
\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}}} \\
\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducestoc \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
\perform{(\outl{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c} \\
\perform{(\outr{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
\end{align*}
\caption{Reduction relations}
\label{fig:reduction}
\end{figure}
\begin{figure}
\begin{align*}
  \mathrm{neutral\ values}, \eta :=\;& x \mid \fst \eta \mid \snd \eta
                                     \mid \gunbox{x}{\eta}{\omega} \mid \grun{\mu} \\
  \mathrm{neutral\ computations}, \mu :=\;& \letv{x}{\mu}{\gamma}
                                           \mid \app{\eta}{\omega} \mid \\
                                   & \case{\eta}{x}{\gamma}{x}{\gamma} \mid \\
                            & \indelim{\Sigma}{\eta}
                              {\indretcase{x}{\gamma} \seq
                               \indcase{op}{x}{x}{\gamma} \seq \ldots \seq
                               \indcase{op}{x}{x}{\gamma}} \\
                                   & \reify{\Sigma}{x}{\mu} \\
  \mathrm{normal\ values}, \omega :=\;& \eta \mid (\omega, \omega)
                           \mid () \mid \inl{\omega} \mid \inr{\omega} \mid \abs{x}{\tau}{\gamma} \mid \\
                           &\indintro{\Sigma}{op}{\omega}{\omega} \mid \ret{\Sigma}{\omega} \mid \\
                           &\gbox{\omega} \mid \outl{\omega} \mid \outr{\omega} \\
  \mathrm{normal\ computations}, \gamma :=\;& \mu \mid \return{\omega}
                                               \mid \perform{\eta}{op}{\omega}{x}{\gamma}
\end{align*}
\caption{Normal forms}
\label{fig:normal}
\end{figure}

@@

\section{Presheaves for algebraic effects}
\label{sec:semantics}

The design of effect reflection falls out naturally from a simple
denotational semantics built on presheaves. This semantic foundation not
only validates our approach but also makes the connection between
locality and effects precise. The key idea is to index the meaning of
types by an ambient algebraic monad of effects: rather than have the
semantics of each type be a set of elements, the semantics of each type
is a mapping from the ambient monad to a set of elements.

@@

As such, we use
\begin{equation*}
  \pshlawcop = \setc^{\lawc}
\end{equation*}
as the basis of our semantics. The $\op$ in the definition of presheaves
and the $\op$ due to our dependence being contravariant cancel out, but
it is still best to think in terms of presheaves as that is a more
precise description of what we are doing.

Our semantics will use objects from $\pshlawcop$ to represent types and
contexts, and morphisms in $\pshlawcop$ to represent values. This allows
their meanings to be parameterized by the ambient monad. By using
presheaves rather than ordinary functions we ensure terms and types at a
particular ambient monad can always be mapped naturally to a larger
ambient monad.

@@

$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
$\pshlawcop$ such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We call $\mon$ the \emph{free algebras monad}, and we will use it as the
computation monad in our denotational semantics.
\begin{align*}
  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
  &\sem{\Gamma \turnc \return{v} \types \tau} =
    \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
    \mu_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ t_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
\end{align*}
where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
respectively of $\mon$.

@@

Within our semantics, this amounts to restricting a type to only the
elements which do not depend on the ambient monad. This is very similar
to the purity comonad of Choudhury and
Krishnaswami~\cite{choudhury2020recovering}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

@@

We use these initial algebras as the denotational semantics of
$\ind{\Sigma}$:
\begin{mathpar}
\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
\end{mathpar}

@@

\appendix
\section{Operational semantics}
\subsection{Subject reduction}
\label{apdx:subject-reduction}
\subsection{Strong normalization}
\label{apdx:strong-normalization}
\section{Denotational semantics}
\label{apdx:denotational}
\subsection{Soundness}
\subsection{Contextual equivalence}
\subsection{Compositionality}
\subsection{Adequacy}
\end{document}

>>>>

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}

  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
  $\glob \tau$, $\yoneda{\Sigma}$ is a type representing how to inject
  terms of $\Sigma$ into the current ambient monad, and $\glob$ is a
  comonadic \emph{global modality} used to track which values do not
  depend on the ambient monad. $\ind{\Sigma}$ is sometimes called the
  \emph{freer monad} of $\Sigma$. This isomorphism emerges naturally
  from the \emph{ambient monad model}, a novel semantic model of
  algebraic effect handlers in terms of presheaves over bounded Lawvere
  theories.

@@

  let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
    let rec loop (s : int) : 'a Term(State).t @ local -> 'a  = function
      | Return x -> x
      | Op(Get, k) -> loop s (State_eff.continue_local k s)
      | Op(Set s', k) -> loop s' (State_eff.continue_local k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
  let get (h : Handler(State).t @ local) : int =
    State_eff.perform Get h

@@

\subsection{The ambient monad model}

This design emerges naturally from a semantic model of algenbraic
effects where types are interpreted as presheaves over \emph{bounded
  Lawvere theories} -- a categorical framework that makes the connection
between locality and effects precise.

@@

The key constructs of effect reflection are then defined as follows:
\begin{itemize}
\item Computations can be represented as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebras.

@@

\item The effect reflection isomorphism becomes a simple consequence of
  the Yoneda lemma and a key property of the sums of algebraic theories.
\end{itemize}

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors. We start with a simple module type
\lstinline[style=oxcaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=oxcaml]
  module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
and \lstinline[style=oxcaml]{Handler} types.
\begin{lstlisting}[style=oxcaml]
  module Continuation(O : Op) : sig
    type ('a, 'b) t
  end

  module Term (O : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r O.t @@ global * ('r, 'a) Continuation(O).t -> 'a t
  end

  module Handler(O : Op) : sig
    type t
  end
\end{lstlisting}
These types directly implement the types from our core isomorphism.
\lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
$\ind{\Sigma}(\glob \tau)$ and represents terms in the free algebra of
\lstinline[style=oxcaml]{Sigma}. \lstinline[style=oxcaml]{Handler(Sigma).t}
corresponds to $\yoneda{\Sigma}$ and represents algebraic effect
handlers handling \lstinline[style=oxcaml]{Sigma} in the current
computation.

\lstinline[style=oxcaml]{('a, 'b) Continuation(Sigma).t} is equivalent
to \lstinline[style=oxcaml]{'a -> 'b Term(Sigma).t} when global and
equivalent to \lstinline[style=oxcaml]{'a -> 'b Term(Sigma).t @ local}
when local. This ensures that the global modality distributes over
\lstinline[style=oxcaml]{Term(Sigma).t}, which is an important property
of $\ind{\Sigma}$. The need for it is an artifact of OxCaml's approach
to locality: the global modality does not distribute over the function
arrow in OxCaml. In the language we describe in
Section~\ref{sec:language} it is replaced with an ordinary function
type.

@@

    val continue : ('a, 'b) Continuation(O).t -> 'a -> 'b Term(O).t

@@

    val continue_local :
      ('a, 'b) Continuation(O).t @ local -> 'a -> 'b Term(O).t @ local

    val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
  end
\end{lstlisting}
\lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_local} implement the left-to-right
direction of the isomorphism. We need two separate functions because
OxCaml lacks any form of mode polymorphism.

\lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
direction of the isomorphism but is a more convenient primitive in
practice. It amounts to reflection of individual operations as opposed
to full terms. We can implement reflection of global full terms as:
\begin{lstlisting}[style=oxcaml]
  let rec reflect t h =
    match t with
    | Return x -> x
    | Op(op, k) -> reflect (continue_local k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
  val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
\end{lstlisting}
Reflection of local terms can be defined similarly.

@@

  let handle_state i f =
    let rec loop (s : int) = function
      | Return x -> x
      | Op(Get, k) -> loop s (State_eff.continue_local k s)
      | Op(Set s', k) -> loop s' (State_eff.continue_localk ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=oxcaml]
  type handler = Handler(State).t

@@

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented directly in OxCaml using
mutable state and locality:
\begin{lstlisting}[style=oxcaml]
  type handler = int ref

  let handle_state i f = f (ref i)
  let get r = !r
  let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state
monad. This shows that the improved reasoning ability of the state monad
over ordinary mutable state is precisely captured by the notion of
locality.

This transformation from effect reflection to local mutable state works
for any \emph{tail-resumptive}\cite{??} effect handler.

@@

  let handle_gen f =
    let rec loop = function
      | Return () -> Finished
      | Op(Gen i, k) -> More(i, fun () -> loop (Gen_eff.continue k ()))
    in
    loop (Gen_eff.reify f)
\end{lstlisting}

@@

  let handle_await f =
    let rec loop = function
      | Return x -> Deferred.return x
      | Op(Await d, k) ->
          Deferred.bind d (fun x -> loop (Await_eff.continue k x))
    in
    loop (Await_eff.reify f)

@@

A key feature of algebraic effects is how easily they compose. We can
use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ah ->
    handle_state i (fun sh ->
      List.iter (fun fl ->
        let s = await ah (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh)) files;
      get sh))
\end{lstlisting}

@@

We could change the definition of \lstinline[style=oxcaml]{handle_gen}
to use \lstinline[style=oxcaml]{reify_local} instead of
\lstinline[style=oxcaml]{reify}, but then we would lose the property
that the generator we produce can be passed around freely. Instead we
need to build a handler that can \emph{forward} the
\lstinline[style=oxcaml]{Await} operation to different effect handlers
at different points in the program's execution.

@@

Composing handlers like \lstinline[style=oxcaml]{handle_gen}, that use
\lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_local}, is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=oxcaml]{Await} operations, so whenever we force part of
the stream we'll need to provide a handler for those
\lstinline[style=oxcaml]{Await} operations again.

@@

  let handle_gen_in_await f ah =
    let rec loop ah = function
      | Return () -> Finished
      | Op(Left op, k) -> loop ah (Eff.continue k (perform ah op))
      | Op(Right(Gen i), k) ->
          More(i, fun ah -> loop ah (Eff.continue k ()))
    in
    loop ah (Eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))
\end{lstlisting}

The need for effect forwarding is a general short-coming of lexical
handlers that applies to all such systems. Examples like this
fundamentally require that performing an \lstinline[style=oxcaml]{Await}
operation does work proportional to the number of handlers it is
forwarded through. Lexical handlers do a constant amount of work to
perform operations and so cannot handle such examples
directly. Non-lexical effect handlers *always* do work proportional to
the number of nested handlers, allowing them to support such examples
directly but paying a higher cost for the cases that do not need
forwarding.

@@

\begin{figure}
\begin{align*}
\fst{(v_1, v_2)} &\reducestov v_1 \\
\snd{(v_1, v_2)} &\reducestov v_2 \\
\gunbox{x}{\gbox{v_1}}{v_2} &\reducestov \sbst{v_2}{x}{v_1} \\
\grun{\return{v}} &\reducestov v \\
\letv{x}{\return{v}}{c} &\reducestoc \sbst{c}{x}{v} \\
\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestoc \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
\app{(\abs{x}{\tau}{c})}{v} &\reducestoc \sbst{c}{x}{v} \\
\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v} \\
\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v} \\
\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducestoc \sbst{c}{x}{v} \\
\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
  \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
\reify{\Sigma}{x}{\return{v}} &\reducestoc \return{\ret{\Sigma}{v}} \\
\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}}} \\
\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducestoc \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
\perform{(\outl{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c} \\
\perform{(\outr{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
\end{align*}
\caption{Reduction relations}
\label{fig:reduction}
\end{figure}
\begin{figure}
\begin{align*}
  \mathrm{neutral\ values}, \eta :=\;& x \mid \fst \eta \mid \snd \eta
                                     \mid \gunbox{x}{\eta}{\omega} \mid \grun{\mu} \\
  \mathrm{neutral\ computations}, \mu :=\;& \letv{x}{\mu}{\gamma}
                                           \mid \app{\eta}{\omega} \mid \\
                                   & \case{\eta}{x}{\gamma}{x}{\gamma} \mid \\
                            & \indelim{\Sigma}{\eta}
                              {\indretcase{x}{\gamma} \seq
                               \indcase{op}{x}{x}{\gamma} \seq \ldots \seq
                               \indcase{op}{x}{x}{\gamma}} \\
                                   & \reify{\Sigma}{x}{\mu} \\
  \mathrm{normal\ values}, \omega :=\;& \eta \mid (\omega, \omega)
                           \mid () \mid \inl{\omega} \mid \inr{\omega} \mid \abs{x}{\tau}{\gamma} \mid \\
                           &\indintro{\Sigma}{op}{\omega}{\omega} \mid \ret{\Sigma}{\omega} \mid \\
                           &\gbox{\omega} \mid \outl{\omega} \mid \outr{\omega} \\
  \mathrm{normal\ computations}, \gamma :=\;& \mu \mid \return{\omega}
                                               \mid \perform{\eta}{op}{\omega}{x}{\gamma}
\end{align*}
\caption{Normal forms}
\label{fig:normal}
\end{figure}

@@

\section{The ambient monad model}
\label{sec:semantics}

The design of effect reflection falls out naturally from a novel
semantic model built on presheaves. This semantic foundation not only
validates our approach but also makes the connection between locality
and effects precise. The key idea is to index the meaning of types by an
ambient algebraic monad of effects: rather than have the semantics of
each type be a set of elements, the semantics of each type is a mapping
from the ambient monad to a set of elements. We give a denotational
semantics to the language from Section~\ref{sec:language} using this
model.

@@

As such, we use
\begin{equation*}
  \pshlawcop = \setc^{\lawc}
\end{equation*}
as the basis of the ambient monad model. The $\op$ in the definition of
presheaves and the $\op$ due to our dependence being contravariant
cancel out, but it is still best to think in terms of presheaves as that
is a more precise description of what we are doing.

The ambient monad model uses objects from $\pshlawcop$ to represent
types and contexts, and morphisms in $\pshlawcop$ to represent
values. This allows their meanings to be parameterized by the ambient
monad. By using presheaves rather than ordinary functions we ensure
terms and types at a particular ambient monad can always be mapped
naturally to a larger ambient monad.

@@

$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
$\pshlawcop$ such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We call $\mon$ the \emph{free algebras monad}, and it is the
representation of computations in the ambient monad model.
\begin{align*}
  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
  &\sem{\Gamma \turnc \return{v} \types \tau} =
    \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
    \mu_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ t_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
\end{align*}
where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
respectively of $\mon$.

@@

Within the abient monad model, this amounts to restricting a type to
only the elements which do not depend on the ambient monad. This is very
similar to the purity comonad of Choudhury and
Krishnaswami~\cite{choudhury2020recovering}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

@@

We use these initial algebras as the semantics of $\ind{\Sigma}$:
\begin{mathpar}
\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
\end{mathpar}

@@

\appendix
\section{Operational semantics}
\subsection{Subject reduction}
\label{apdx:subject-reduction}

We prove subject reduction (Lemma~\ref{lemma:subject-reduction}):
\begin{equation*}
  \Gamma \turnv v_1 \types \tau
  \wedge v_1 \reducestov v_2 \implies \Gamma \turnv v_2 \types \tau
\end{equation*}
by mutual induction over the two reduction relations. The reduction
relations are the least reflexive, transitive congruences closed under
the rules in Fig.~\ref{fig:reduction}.

The reflexivity case is immediate and the transitive case follows
directly by induction. As all the typing rules are entirely
compositional, the congruence cases also all follow direectly by
induction. This leaves only the rules from Fig.~\ref{fig:reduction}.

Note that the typing rules are syntax directed with the exception of the
two var rules, which are distinguished by how the variable is bound in
the context. Since none of the left-hand sides of our reduction rules is
a variable, there is always only a single typing rule that we need to
consider.

The cases for the products rules:
\begin{align*}
\fst{(v_1, v_2)} &\reducestov v_1 \\
\snd{(v_1, v_2)} &\reducestov v_2 \\
\end{align*}
follow immediately by inversion on the typing rule.

\begin{equation*}
\gunbox{x}{\gbox{v_1}}{v_2} \reducestov \sbst{v_2}{x}{v_1}
\end{equation*}
\todo{Requires substitution rule of a term under $\cdiv{\Gamma}$ for a
  boxed binding}.

The case for the $\grun$ rule
\begin{equation*}
\grun{\return{v}} \reducestov v
\end{equation*}
\todo{Requires weakening from $\cdiv{\Gamma}$ to $\Gamma$}.

The case for the let-return rule:
\begin{equation*}
\letv{x}{\return{v}}{c} \reducestoc \sbst{c}{x}{v}
\end{equation*}
\todo{Requires ordinary substitution lemma}.

The case for the let-perform rule:
\begin{equation*}
\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} \reducestoc \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}}
\end{equation*}
follows immediately by inversion on the typing rule.

The case for the standard beta rule:
\begin{equation*}
\app{(\abs{x}{\tau}{c})}{v} \reducestoc \sbst{c}{x}{v}
\end{equation*}
\todo{Requires ordinary substitution lemma}.

The cases for the case rules:
\begin{align*}
\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v}
\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v}
\end{align*}
\todo{Requires ordinary substitution lemma}.

The case for the $\retname$ rule:
\begin{equation*}
\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} \reducestoc \sbst{c}{x}{v}
\end{equation*}
\todo{Requires ordinary substitution lemma}.
